---
title: "Using fdaMixedRTMB: boundary-value models, diagnostics, and marginal means"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using fdaMixedRTMB: boundary-value models, diagnostics, and marginal means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

`fdaMixedRTMB` fits operator-based functional mixed models where each
observed curve is represented on a common time grid. This vignette uses the
packaged beer-tracker data to show the typical workflow:

1. We turn event-level records into cumulative and weekly trajectory data.
2. Fit a boundary-value model with `fdaLm_rtmb()`.
3. Inspect the fitted object with package methods.
4. Use diagnostic plots.
5. Use `emmeans` for the boundary-value regression.
6. Extract fitted curves for plotting or downstream summaries.

```{r libraries}
library(fdaMixedRTMB)
library(emmeans)
library(tidyverse)
library(lubridate)

old_theme <- theme_set(theme_minimal(base_size = 11))

sex_cols <- c(F = "#D55E00", M = "#0072B2")
```

## Prepare Beer-Tracker Data

The example data are event records: one row per logged beer. A functional
response needs one row per subject and time point, so the first step is to
aggregate to daily counts and fill missing subject-day combinations with zero.

```{r read-data}
beer_file <- system.file("examples", "dailybeers.csv", package = "fdaMixedRTMB")

if (!nzchar(beer_file)) {
  beer_file <- file.path("..", "inst", "examples", "dailybeers.csv")
}

sex_lookup <- tibble(
  navn = c(
    "Mille", "Milla", "Augusta", "Lise", "Frigg",
    "Baldur", "Julle", "Malthe", "Nikolai", "William"
  ),
  sex = c(rep("F", 5), rep("M", 5))
)

beer_raw <- read_csv(beer_file, show_col_types = FALSE) |>
  mutate(
    timestamp = as.POSIXct(
      tid,
      format = "%d/%m/%Y %H.%M.%S",
      tz = "UTC"
    ),
    date = as_date(timestamp),
    beer = str_replace(as.character(antal), ",", "."),
    beer = as.numeric(beer),
    beer = replace_na(beer, 1)
  ) |>
  select(navn, date, beer) |>
  left_join(sex_lookup, by = "navn")

range(beer_raw$date)

beer_raw |>
  count(navn)
```

The names are anonymized before modeling. `design` data frames in
`fdaMixedRTMB` are curve-level data frames: they contain one row per curve and
the covariates that are constant for that curve. The raw log has no beer
entries after May 26, but the observation window runs through May 28, so the
daily grid below explicitly includes May 27 and May 28 as zero-count days.

```{r daily-grid}
last_beer_date <- max(beer_raw$date, na.rm = TRUE)
analysis_year <- year(last_beer_date)

analysis_start <- ymd(paste0(analysis_year, "-01-01"))
analysis_end <- max(
  last_beer_date,
  ymd(paste0(year(last_beer_date), "-", month(last_beer_date), "-28"))
)

months_ordered <- month.name[seq_len(month(analysis_end))]
n_days_month <- min(28L, day(analysis_end))

c(
  last_logged_beer = as.character(last_beer_date),
  analysis_end = as.character(analysis_end)
)

date_grid <- tibble(
  date = seq(analysis_start, analysis_end, by = "day")
) |>
  mutate(
    day = row_number(),
    month_num = month(date),
    month_name = factor(month.name[month_num], levels = month.name),
    day_within_month = day(date)
  )

subject_lookup <- beer_raw |>
  distinct(navn, sex) |>
  arrange(navn) |>
  mutate(
    subject = row_number(),
    name = factor(paste("Friend", subject), levels = paste("Friend", subject)),
    sex = factor(sex, levels = c("F", "M"))
  )

daily_totals <- beer_raw |>
  group_by(navn, sex, date) |>
  summarise(beer = sum(beer), .groups = "drop")

beer_daily <- subject_lookup |>
  crossing(date_grid) |>
  left_join(daily_totals, by = c("navn", "sex", "date")) |>
  mutate(beer = replace_na(beer, 0)) |>
  arrange(subject, day) |>
  group_by(subject) |>
  mutate(beer_gain = cumsum(beer)) |>
  ungroup()

beer_daily |>
  select(name, sex, date, day, beer, beer_gain) |>
  head()

beer_daily |>
  filter(date >= analysis_end - days(1)) |>
  group_by(date) |>
  summarise(beer = sum(beer), .groups = "drop")
```

### Full-Period Curves

Full-period curves accumulate beer gains from the first analysis day through
the last available day. The right endpoint of such a curve is the subject's
total beer gain over the full tracker period.

```{r full-period-curves}
full_period_df <- beer_daily |>
  mutate(
    time = day,
    y = beer_gain,
    curve = name
  )

full_curve_meta <- full_period_df |>
  distinct(curve, subject, name, sex) |>
  arrange(subject)

ggplot(full_period_df, aes(time, y, colour = sex)) +
  geom_line(aes(group = name), alpha = 0.3, linewidth = 0.4) +
  stat_summary(aes(group = sex), fun = mean, geom = "line", linewidth = 1.1) +
  scale_colour_manual(values = sex_cols) +
  labs(
    title = "Full-period cumulative beer-gain curves",
    x = "Day of tracker period",
    y = "Cumulative beers",
    colour = NULL
  )
```

A full-period model uses the same interface. For example, the endpoint
boundary derivative could be modeled by sex:

```{r full-period-fit-example, eval = FALSE}
fit_full <- fdaLm_rtmb(
  y | curve ~ 0,
  data = full_period_df,
  design = full_curve_meta,
  boundary_formula = ~ sex,
  boundary_value = TRUE,
  operator = operator_laplace(),
  right_boundary = right_boundary_operator_k1(alpha = c(0, 1)),
  time_variable = "time"
)
```

### Monthly Curves

Monthly curves reset the cumulative sum at the beginning of each month. To put
all months on the same time grid, this vignette uses the common number of
available days, capped at 28.

```{r monthly-curves}
beer_monthly_df <- beer_daily |>
  filter(
    month_name %in% months_ordered,
    day_within_month <= n_days_month
  ) |>
  arrange(subject, month_num, day_within_month) |>
  group_by(subject, month_num) |>
  mutate(monthly_cumsum = cumsum(beer)) |>
  ungroup() |>
  mutate(
    month_name = factor(as.character(month_name), levels = months_ordered),
    curve = paste(name, month_name, sep = ": ")
  )

curve_meta <- beer_monthly_df |>
  distinct(curve, subject, name, sex, month_name, month_num) |>
  arrange(subject, month_num) |>
  mutate(
    curve = factor(curve, levels = curve),
    name = factor(name, levels = levels(subject_lookup$name)),
    sex = factor(sex, levels = c("F", "M")),
    month_name = factor(month_name, levels = months_ordered)
  )

beer_monthly_df <- beer_monthly_df |>
  mutate(curve = factor(curve, levels = levels(curve_meta$curve)))

beer_monthly_df |>
  select(name, month_name, day_within_month, beer, monthly_cumsum, curve) |>
  head()
```

```{r monthly-plot}
monthly_means <- beer_monthly_df |>
  group_by(month_name, day_within_month) |>
  summarise(monthly_cumsum = mean(monthly_cumsum), .groups = "drop")

month_cols <- setNames(
  grDevices::hcl.colors(length(months_ordered), "Dark 3"),
  months_ordered
)

ggplot(monthly_means, aes(day_within_month, monthly_cumsum, colour = month_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = month_cols) +
  labs(
    title = "Average monthly beer-gain curves",
    x = "Day within month",
    y = "Cumulative beers",
    colour = NULL
  )
```

## Fit a Boundary-Value Model

The model below uses monthly cumulative beer gains as the response. For a
faster and more readable vignette, the fit uses four anonymized participants;
the data-preparation code above created the full set of monthly curves.

```{r model-data}
model_subjects <- levels(subject_lookup$name)[1:4]

curve_meta_model <- curve_meta |>
  filter(name %in% model_subjects) |>
  mutate(curve = factor(as.character(curve), levels = as.character(curve)))

beer_monthly_model <- beer_monthly_df |>
  filter(name %in% model_subjects) |>
  mutate(curve = factor(as.character(curve), levels = levels(curve_meta_model$curve)))

c(
  observations = nrow(beer_monthly_model),
  curves = nlevels(beer_monthly_model$curve),
  days_per_curve = n_days_month
)
```

```{r fit-model}
fit <- fdaLm_rtmb(
  monthly_cumsum | curve ~ 0,
  data = beer_monthly_model,
  design = curve_meta_model,
  boundary_formula = ~ month_name,
  boundary_value = TRUE,
  operator = operator_laplace(),
  right_boundary = right_boundary_operator_k1(alpha = c(0, 1)),
  time_variable = "day_within_month"
)
```

The main arguments are:

- `monthly_cumsum | curve ~ 0`: the response is `monthly_cumsum`, and `curve`
  identifies the functional curve. The fixed-effect part is `0` because this
  example puts the month effect in the boundary-value regression.
- `boundary_value = TRUE`: estimate a nonzero right-boundary value.
- `boundary_formula = ~ month_name`: model the curve-level boundary value as a
  function of month.
- `right_boundary = right_boundary_operator_k1(alpha = c(0, 1))`: define the
  right boundary operator as endpoint derivative, so the fitted `rho` is the
  month-end derivative.
- `operator = operator_laplace()`: use the K.order = 1 pure Laplace operator
  `L f = -lambda_d2 f''`. The shifted operator is available via
  `operator_k1(lambda_start = c(1, 1))`.

## Inspect the Fitted Model

The fitted object has ordinary R methods for printing, summarizing,
coefficients, predictions, residuals, log likelihoods, and model dimensions.

```{r print-summary}
fit
summary(fit)
```

```{r coefficients}
coef(fit, component = "boundary")
coef(fit, component = "rho")
coef(fit, component = "variance")
```

`component = "boundary"` gives the coefficients of the boundary-value
regression. `component = "rho"` returns the fitted curve-specific boundary
values.

## Plot Diagnostics

The S3 diagnostic method is available as
`plot(fit, which = c("fit", "residual", "rho"))`. The vignette below builds
the same common diagnostics with `ggplot2` so the plot style is consistent:

- `"fit"`: observed values against fitted values,
- `"residual"`: residuals against fitted values, and
- `"rho"`: estimated curve-specific right-boundary values.

```{r diagnostics}
diagnostic_df <- tibble(
  observed = fit$data$y,
  fitted = as.numeric(predict(fit, type = "response")),
  residual = as.numeric(residuals(fit))
)

rho_df <- tibble(
  curve = names(coef(fit, component = "rho")),
  rho = as.numeric(coef(fit, component = "rho"))
) |>
  mutate(curve = factor(curve, levels = curve))

ggplot(diagnostic_df, aes(fitted, observed)) +
  geom_point(alpha = 0.65, size = 1.2) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40") +
  labs(
    title = "Observed vs fitted",
    x = "Fitted value",
    y = "Observed value"
  )

ggplot(diagnostic_df, aes(fitted, residual)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_point(alpha = 0.65, size = 1.2) +
  labs(
    title = "Residuals vs fitted",
    x = "Fitted value",
    y = "Residual"
  )

ggplot(rho_df, aes(curve, rho)) +
  geom_col(fill = "grey45") +
  coord_flip() +
  labs(
    title = "Estimated boundary values",
    x = "Curve",
    y = "Estimated rho"
  )
```

The `"boundary"` panel is also useful: it checks the right boundary operator on
the homogeneous serial effect. For a well-behaved boundary-value decomposition,
this quantity should be close to zero relative to the scale of `rho`.

```{r boundary-check}
boundary_check_df <- tibble(
  curve = names(fit$right_boundary_serial_operator),
  right_operator = as.numeric(fit$right_boundary_serial_operator)
) |>
  mutate(curve = factor(curve, levels = curve))

ggplot(boundary_check_df, aes(curve, right_operator)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_point(size = 1.6) +
  coord_flip() +
  labs(
    title = "Homogeneous boundary check",
    x = "Curve",
    y = "Right operator on serial effect"
  )
```

## Use emmeans

`fdaMixedRTMB` provides `emmeans` support for both ordinary fixed effects and
the boundary-value regression. Here we ask for marginal means of `month_name`
in `mode = "boundary"`, which targets the fitted `rho ~ month_name` model.

```{r emmeans}
emm_month <- emmeans(fit, ~ month_name, mode = "boundary")
emm_month
pairs(emm_month)
```

Because the right boundary is `alpha = c(0, 1)`, these marginal means are
estimated endpoint derivatives. If the right boundary were `alpha = c(1, 0)`,
the same `emmeans()` call would summarize the month-end cumulative beer gains
instead.

## Extract Fitted Curves

Use `predict()` and `residuals()` to recover fitted values and model
components on the original curve grid. The returned vectors follow the internal
curve-time ordering, so the data are sorted by `curve` and time before adding
them.

```{r extract-fitted}
fitted_curves <- beer_monthly_model |>
  arrange(curve, day_within_month) |>
  mutate(
    fitted = as.numeric(predict(fit, type = "response")),
    boundary_effect = as.numeric(predict(fit, type = "boundary")),
    serial_effect = as.numeric(predict(fit, type = "serial")),
    residual = as.numeric(residuals(fit))
  )

fitted_curves |>
  select(
    name, month_name, day_within_month, monthly_cumsum,
    fitted, boundary_effect, serial_effect, residual
  ) |>
  head()
```

```{r fitted-plot}
fitted_month_means <- fitted_curves |>
  group_by(month_name, day_within_month) |>
  summarise(fitted = mean(fitted), .groups = "drop")

ggplot(fitted_month_means, aes(day_within_month, fitted, colour = month_name)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = month_cols) +
  labs(
    title = "Fitted monthly beer-gain curves",
    x = "Day within month",
    y = "Fitted cumulative beers",
    colour = NULL
  )
```

## Weekly Trajectories Without Cumulative Sums

The cumulative curves are useful when the scientific question is about total
gain by the end of a period. A different question is whether the weekly
numbers themselves change over time. In that case each curve is one person's
tracker-week beer totals, the time grid is week number, and the response is
`weekly_beers` rather than an accumulated total. Weeks are consecutive 7-day
blocks anchored at January 1. The observation window ends on May 28, which
creates one trailing zero-only day; that day is folded into the final tracker
week so the last plotted point is the May 21-28 week rather than an artificial
zero-only week.

For non-cumulative weekly counts, the endpoint value is just the beer count in
the final modeled week. The endpoint derivative is more useful: it estimates
the local trend at the end of the observed period, in weekly beers per week.
This example therefore uses `right_boundary = c(0, 1)` and a boundary
regression `rho ~ sex`.

```{r weekly-curves}
beer_weekly_daily <- beer_daily |>
  mutate(raw_week_num = ((day - 1L) %/% 7L) + 1L)

last_raw_week <- max(beer_weekly_daily$raw_week_num)

last_raw_week_total <- beer_weekly_daily |>
  filter(raw_week_num == last_raw_week) |>
  summarise(total = sum(beer), .groups = "drop") |>
  pull(total)

beer_weekly_daily <- beer_weekly_daily |>
  mutate(
    week_num = if_else(
      last_raw_week_total == 0 & last_raw_week > 1L & raw_week_num == last_raw_week,
      last_raw_week - 1L,
      raw_week_num
    )
  )

n_weeks <- max(beer_weekly_daily$week_num)

beer_weekly_df <- beer_weekly_daily |>
  group_by(subject, name, sex, week_num) |>
  summarise(weekly_beers = sum(beer), .groups = "drop") |>
  arrange(subject, week_num) |>
  mutate(
    week_index = week_num - 1L,
    week_label = factor(
      paste("Week", week_index),
      levels = paste("Week", seq_len(n_weeks) - 1L)
    ),
    curve = name
  )

weekly_curve_meta <- beer_weekly_df |>
  distinct(curve, subject, name, sex) |>
  arrange(subject) |>
  mutate(
    curve = factor(curve, levels = curve),
    name = factor(name, levels = levels(subject_lookup$name)),
    sex = factor(sex, levels = c("F", "M"))
  )

beer_weekly_df <- beer_weekly_df |>
  mutate(curve = factor(curve, levels = levels(weekly_curve_meta$curve)))

beer_weekly_df |>
  select(name, sex, week_index, week_label, weekly_beers, curve) |>
  head()
```

```{r weekly-plot}
ggplot(beer_weekly_df, aes(week_index, weekly_beers, colour = sex)) +
  geom_line(aes(group = name), alpha = 0.25, linewidth = 0.4) +
  geom_point(aes(group = name), alpha = 0.35, size = 0.8) +
  stat_summary(aes(group = sex), fun = mean, geom = "line", linewidth = 1.25) +
  scale_colour_manual(values = sex_cols) +
  scale_x_continuous(breaks = seq(0, n_weeks - 1L, by = 5)) +
  labs(
    title = "Weekly beers",
    x = "Week number",
    y = "Weekly beers",
    colour = NULL
  )
```

```{r weekly-fit}
fit_weekly <- fdaLm_rtmb(
  weekly_beers | curve ~ 0,
  data = beer_weekly_df,
  design = weekly_curve_meta,
  boundary_formula = ~ sex,
  boundary_value = TRUE,
  operator = operator_laplace(),
  right_boundary = right_boundary_operator_k1(alpha = c(0, 1)),
  time_variable = "week_index"
)

c(
  observations = nrow(beer_weekly_df),
  curves = nlevels(beer_weekly_df$curve),
  weeks_per_curve = n_distinct(beer_weekly_df$week_num),
  converged = fit_weekly$converged
)
```

The boundary coefficients and marginal means now describe sex differences in
the right-end derivative of the weekly-number trajectory. Because the right
boundary is `alpha = c(0, 1)`, `rho` is the fitted endpoint slope rather than
the fitted number of beers in the final week.

```{r weekly-emmeans}
coef(fit_weekly, component = "boundary")
coef(fit_weekly, component = "rho")
coef(fit_weekly, component = "variance")

emm_weekly_slope <- emmeans(fit_weekly, ~ sex, mode = "boundary")
emm_weekly_slope
pairs(emm_weekly_slope)
```

```{r weekly-slope-plot}
weekly_slope_df <- as_tibble(emm_weekly_slope)

ggplot(weekly_slope_df, aes(sex, emmean, colour = sex)) +
  geom_hline(yintercept = 0, colour = "grey45") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.12) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = sex_cols, guide = "none") +
  labs(
    title = "Estimated endpoint derivative of weekly beers",
    x = NULL,
    y = "Endpoint slope, weekly beers per week"
  )
```

The Gaussian model can produce fitted values below zero for sparse count data,
so this fit should be read as a functional-trajectory demonstration rather than
as a full count model.

```{r weekly-fitted}
weekly_fitted_curves <- beer_weekly_df |>
  arrange(curve, week_index) |>
  mutate(
    fitted = as.numeric(predict(fit_weekly, type = "response")),
    residual = as.numeric(residuals(fit_weekly))
  )

weekly_mean_observed <- weekly_fitted_curves |>
  group_by(sex, week_index) |>
  summarise(value = mean(weekly_beers), .groups = "drop") |>
  mutate(type = "Observed mean")

weekly_mean_fitted <- weekly_fitted_curves |>
  group_by(sex, week_index) |>
  summarise(value = mean(fitted), .groups = "drop") |>
  mutate(type = "Fitted mean")

weekly_mean_plot <- bind_rows(weekly_mean_observed, weekly_mean_fitted)

ggplot(weekly_mean_plot, aes(week_index, value, colour = sex, linetype = type)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = sex_cols) +
  scale_x_continuous(breaks = seq(0, n_weeks - 1L, by = 5)) +
  labs(
    title = "Fitted weekly beer trajectories",
    x = "Week number",
    y = "Weekly beers",
    colour = NULL,
    linetype = NULL
  )
```

## Known Limitations

At the time of writing, user-specified grouped random effects are not part of
the public modeling workflow. The practical supported pattern is to use the
formula interface for ordinary fixed effects and to use `boundary_formula` for
curve-level boundary-value regressions. The serial functional effect is handled
internally through the operator-projection machinery.